---
title: "LMD Study Analysis"
author: "Quantitative Sciences Unit, Stanford Medicine <br> Bo Gu"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
    toc_depth: 4
    theme: yeti
    highlight: tango
    code_folding: hide
  pdf_document:
    toc: yes
    toc_depth: '4'
---

```{=html}
<style type="text/css">
body .main-container {
max-width: 1700px;
}

h1.title {
  font-size: 32px;
  color: DarkRed;
  text-align: center;
}

h3.subtitle {
  font-size: 22px;
  color: DarkRed;
  text-align: center;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  color: Black;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  color: Black;
  text-align: center;
}
caption {
  color: Black;
  font-weight: bold;
  font-size: 1.2em;
} 
th, td {
    padding-left: 10px;
    font-size: 1.2em;
    text-align: center;        
}
</style>
```

```{r, echo = T, message = F, warning = F}
library(survival)
library(readr)
library(readxl)
library(lubridate)
library(table1)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(survminer)
library(car)
library(MASS)
library(stargazer)
library(tidyr)
library(olsrr)
library(mice)
library(miceadds)


LMD_cohort <- read_csv('/Users/bgu/Library/CloudStorage/Box-Box/LMD-HR-breast cancer Sophia/Data/(Updated 062623) Full Data Sheet.csv') %>% 
  as.data.frame()


LMD_cohort <- LMD_cohort  %>% 
  mutate(Birth.date = ymd(`Date of Birth`), 
         Death.date = ymd(`Date of Death`), 
         Lfup.date = ymd(`Date of Last Contact`), 
         Surv.date = case_when(is.na(Death.date) ~ Lfup.date, 
                               T ~ Death.date),
         Brc.date = ymd(`Diagnosis Date`), 
         First.met.date = ymd(`1st Met Date`), 
         Brain.met.date = ymd(`Brain Met Date`), 
         LMD.date = ymd(`LMD Met Date`), 
         Status = `Censor Status`, 
         ER = as.numeric(ER), 
         PR = as.numeric(PR), 
         HER2 = as.numeric(HER2),
         Stage = factor(Stage, levels = c(1, 2, 3, 4))) %>% 
  mutate(Age = round(interval(Birth.date, Brc.date) / years(1), 0), 
         Race = as.factor(ifelse(Race == 'Other' | Race == 'Pacific Islander' | Race == 'Unknown', 'Other', Race)),
         Brc.first.met = interval(Brc.date, First.met.date) %/% days(1) / (365/12), 
         Brc.brain = interval(Brc.date, Brain.met.date) %/% days(1) / (365/12), 
         Brc.LMD = interval(Brc.date, LMD.date) %/% days(1) / (365/12),
         First.brain = interval(First.met.date, Brain.met.date) %/% days(1) / (365/12),
         First.LMD = interval(First.met.date, LMD.date) %/% days(1) / (365/12),
         Brain.LMD = interval(Brain.met.date, LMD.date) %/% days(1) / (365/12),
         LMD.surv = interval(LMD.date, Surv.date) %/% days(1) / (365/12)) %>% 
  filter(!is.na(Brc.LMD)) %>% 
  mutate(First.brain.LMD = case_when(is.na(Brc.first.met) & !is.na(Brc.brain) & Brc.brain != Brc.LMD ~ 'First Met in Brain', 
                                     Brc.first.met == Brc.brain & Brc.brain != Brc.LMD ~ 'First Met in Brain',
                                     is.na(Brc.first.met) & is.na(Brc.brain) ~ 'First Met is LMD', 
                                     Brc.first.met == Brc.LMD & is.na(Brc.brain) ~ 'First Met is LMD',
                                     Brc.first.met == Brc.LMD & Brc.first.met == Brc.brain ~ 'First Met in Brain and is LMD',
                                     is.na(Brc.first.met) & Brc.brain == Brc.LMD  ~ 'First Met in Brain and is LMD',
                                     T ~ 'First Met not in Brain / not LMD'), 
         First.met.bi = case_when(`1st Met Site` == 'Brain' ~ 'Brain', 
                                  !grepl('Brain|LMD', `1st Met Site`, ignore.case = T) ~ 'Non-Brain'), 
         First.met.bi.LMD = as.factor(case_when(grepl('Brain|LMD', `1st Met Site`, ignore.case = T) ~ 'Brain', 
                                      T ~ 'Non-Brain')),
         ER = factor(1 - ER, labels = c('Positive', 'Negative')), 
         PR = factor(1 - PR, labels = c('Positive', 'Negative')), 
         HER2 = factor(1 - HER2, labels = c('Positive', 'Negative')))


HD_corhort <- read_csv('/Users/bgu/Library/CloudStorage/Box-Box/LMD-HR-breast cancer Sophia/Data/(070323) HD Full Table.csv') %>%
  fill(`Starr ID`, .direction = 'down') %>%
  mutate(`HR Date` = as.Date(Date),
         ER = factor(1 - as.numeric(ER), labels = c('Positive', 'Negative')),
         PR = factor(1 - as.numeric(PR), labels = c('Positive', 'Negative')),
         HER2 = factor(1 - as.numeric(HER2), labels = c('Positive', 'Negative'))) %>%
  dplyr::select(-Date) %>%
  arrange(`Starr ID`, `HR Date`) %>%
  distinct(`Starr ID`, ER, PR, HER2, `HR Date`, .keep_all = T)


LMD_long <- HD_corhort %>% 
  dplyr::select(-Location) %>% 
  left_join(LMD_cohort %>% 
              mutate(`HR Date` = mdy(`HR Date`)) %>% 
              dplyr::select(1:2, 17, 19:22, 23, 24, 44:57, 60), by = 'Starr ID') %>% 
  group_by(`Starr ID`) %>% 
  mutate(ID = cur_group_id(),
         Subtype = as.factor(tolower(Subtype)),
         `Stage (T)` = as.factor(tolower(`Stage (T)`)),
         `Stage (N)` = as.factor(tolower(`Stage (N)`)),
         `Stage (M)` = as.factor(tolower(`Stage (M)`)),
         Grade = as.factor(tolower(Grade)),
         `1st Met Site` = as.factor(tolower(`1st Met Site`))) %>% 
  arrange(`Starr ID`, `HR Date`) %>% 
  rename(Starr.ID = `Starr ID`,
         Stage.T = `Stage (T)`,
         Stage.N = `Stage (N)`,
         Stage.M = `Stage (M)`,
         First.met.site = `1st Met Site`, 
         HR.Date = `HR Date`) %>% 
  ungroup() %>% 
  dplyr::select(ID, Race, Age, Subtype, Stage, Grade, Status, HR.Date, Brc.date, Brc.first.met, Brc.LMD, LMD.surv, First.met.bi.LMD, ER, PR, HER2) %>% 
  filter(!is.na(HR.Date)) %>% 
  group_by(ID) %>% 
  mutate(HR.Date = case_when(row_number() == 1 & HR.Date != Brc.date ~ Brc.date,
                             T ~ HR.Date),
         First.met.site = First.met.bi.LMD) %>%
  mutate(Brc.first.met = ifelse(Brc.first.met >= 0, Brc.first.met, 0),
         Brc.LMD = ifelse(Brc.LMD >= 0, Brc.LMD, 0)) %>%
  mutate(Tstart = interval(Brc.date, HR.Date) %/% days(1) / (365/12)) %>% 
  as.data.frame()


pred <- make.predictorMatrix(LMD_long)
pred[, 'HR.Date'] <- 0
pred[, 'Brc.date'] <- 0
pred[, 'ID'][c(5:6, 14:16)] <- -2
pred['ER', 15:16] <- 2
pred['PR', c(14, 16)] <- 2
pred['HER2', c(14:15)] <- 2

meth <- make.method(LMD_long)
meth[c(5:6)] <- "2lonly.pmm"
meth[8] <- ''
meth[c(14:16)] <- "2l.bin"



imp_LMD_long <- mice(LMD_long, m = 20, maxit = 30, pred = pred, meth = meth, seed = 1024, print = F)

complete_LMD_long <- complete(imp_LMD_long, 'long', T)


# 139
base_LMD <- complete_LMD_long %>% 
  group_by(.imp, ID) %>% 
  filter(row_number() == 1) %>% 
  as.data.frame() 

# n_distinct(base_LMD$ID)


# 85
base_LMD_subgroup <- base_LMD %>% 
  filter(First.met.site == 'Non-Brain') %>% 
  as.data.frame()

# n_distinct(base_LMD_subgroup$ID)

# 130
HD_1st_met <- complete_LMD_long %>% 
  group_by(.imp, ID) %>% 
  filter(Brc.first.met > Tstart) %>% 
  mutate(Tstop = lead(Tstart, n = 1L)) %>% 
  mutate(Tstop = ifelse(Brc.first.met < Tstop & row_number() == n() | is.na(Tstop), Brc.first.met, Tstop), 
         Event = case_when(First.met.site == 'Non-Brain' & row_number() == n() ~ 1, 
                           First.met.site == 'Brain' & row_number() == n() ~ 2, 
                           T ~ 0)) %>% 
  # mutate(ER = ifelse(ER == 'Positive', 0, 1)) %>% 
  as.data.frame()

# n_distinct(HD_1st_met$ID)


# 130
base_1st_met <- HD_1st_met %>% 
  group_by(.imp, ID) %>% 
  filter(row_number() == n()) %>% 
  as.data.frame()

# n_distinct(base_1st_met$ID)


# 137
HD_LMD <- complete_LMD_long %>% 
  group_by(.imp, ID) %>% 
  filter(Brc.LMD > Tstart) %>% 
  mutate(Tstop = lead(Tstart, n = 1L)) %>% 
  mutate(Tstop = ifelse(is.na(Tstop), Brc.LMD, Tstop), 
         Event = 1) %>% 
  # mutate(ER = ifelse(ER == 'Positive', 0, 1)) %>% 
  mutate(mark = ifelse(Brc.first.met > Tstart & Brc.first.met < Tstop, 2, 1)) %>% 
  uncount(., mark, .remove = F) %>% 
  group_by(.imp, ID, mark) %>% 
  mutate(Tstart = ifelse(mark == 2 & row_number() == 2, Brc.first.met, Tstart),
         Tstop = ifelse(mark == 2 & row_number() == 1, Brc.first.met, Tstop)) %>% 
  mutate(First.met.brain = as.factor(ifelse(First.met.site == 'Brain' & Tstart >= Brc.first.met, 'Yes', 'No')), 
         First.met.non.brain = as.factor(ifelse(First.met.site == 'Non-Brain' & Tstart >= Brc.first.met, 'Yes', 'No')),
         First.met.state = as.factor(case_when(First.met.brain == 'Yes' ~ 2, 
                                     First.met.non.brain == 'Yes' ~ 1,
                                     T ~ 0))) %>% 
  filter(Tstop != Tstart) %>% 
  group_by(.imp) %>% 
  mutate(.id = row_number()) %>% 
  as.data.frame()


# 137
base_HD_LMD <- HD_LMD %>% 
  group_by(.imp, ID) %>% 
  filter(row_number() == 1) %>% 
  as.data.frame() 

# n_distinct(base_HD_LMD$ID)

base_LMD <- as.mids(base_LMD)

base_LMD_subgroup <- as.mids(base_LMD_subgroup)

HD_1st_met <- as.mids(HD_1st_met)

base_1st_met <- as.mids(base_1st_met)

HD_LMD <- as.mids(HD_LMD)

base_HD_LMD <- as.mids(base_HD_LMD)

  
get_pool_coef <- function(fit){
  sum <- summary(pool(fit), conf.int=T)
  cbind.data.frame('Variables' = sum[, 1], 'Coefficients [95% CI]' = paste0(round(sum[, "estimate"], 1), ' [', round(sum[, "2.5 %"], 1), ', ', round(sum[, "97.5 %"], 1), ']'), 
        'p-value' = round(sum[, 'p.value'], 3))
}


get_pool_coef_cox <- function(fit){
  sum <- summary(pool(fit), conf.int=T)
  cbind.data.frame('Variables' = sum[, 1], 'Coefficients [95% CI]' = paste0(round(exp(sum[, "estimate"]), 1), ' [', round(exp(sum[, "2.5 %"]), 1), ', ', round(exp(sum[, "97.5 %"]), 1), ']'), 
        'p-value' = round(sum[, 'p.value'], 3))
}
```

### 1. Check if there are cases that Brc date is different from HR date

```{r, echo = T, message = F, warning = F}
# Check if there are cases that Brc date is different from HR date
LMD_cohort %>%
  mutate(ind = ifelse(`Diagnosis Date` != `HR Date`, 1, 0)) %>%
  filter(ind == 1) %>%
  dplyr::select(`Starr ID`, `Diagnosis Date`, `HR Date`, `Date of Death`, `Date of Last Contact`, Status) %>%
  kbl(align = 'ccc', row.names = F) %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)
```

### 2. Check if there are cases that last fup date is different from survival date

```{r, echo = T, message = F, warning = F}
# Check if there are cases that last fup date is different from survival date
LMD_cohort %>%
  mutate(ind = ifelse(Lfup.date != Surv.date, 1, 0)) %>%
  filter(ind == 1 | grepl('unconfirmed|deceased', `Date of Death`, ignore.case = T)) %>%
  dplyr::select(`Starr ID`, `Date of Death`, Lfup.date, Surv.date, Status) %>%
  kbl(align = 'ccccc', row.names = F) %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)
```

### 3. Check Age: Negative ages

```{r, echo = T, message = F, warning = F}
LMD_cohort %>%
  filter(Age <= 0) %>%
  dplyr::select(`Starr ID`, `Date of Birth`, `Diagnosis Date`, Age) %>%
  kbl(align = 'ccc', row.names = F) %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)

```

### 4. Check First Met, Brain Met, LMD and their relationship

```{r, echo = T, message = F, warning = F}
# Check First Met, Brain Met, LMD and their relationship
table1(~ Brc.first.met + Brc.brain + Brc.LMD + LMD.surv + First.brain.LMD + `1st Met Site`, data = LMD_cohort, caption = 'Relationship among First Met, Brain Met and LMD')


# Discrepancy between original 1st site and location determined by time
LMD_cohort %>%
  dplyr::select(1, 24, 52:58) %>%
  filter(First.brain.LMD == 'First Met in Brain' & `1st Met Site` != 'Brain') %>%
  kbl(align = 'cccccc', row.names = F) %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)


```

### 5. Demographics

```{r, echo = T, message = F, warning = F}
# Demographics
table1(~ Age + Race + Ethnicity , data = LMD_cohort, caption = 'LMD Cohort: Demographics')
```

### 6. Some covariates of interest

```{r, echo = T, message = F, warning = F}
# Some covariates of interest
table1 (~ ER + PR + HER2 + Stage, data = LMD_cohort, caption = 'LMD Cohort: Covariates')
```

#### 18% of "Stage" are missing, data impuation is needed if we still want to include it.

```{r, echo = T, message = F, warning = F}
tab <- xtabs(~ ER + PR + HER2, data = LMD_cohort, addNA = T)

tab
```

#### According to 3-way table above, there might be a moderate to high correlation between ER and PR.

### 7. KM plots by ER, PR and HER2 status (LMD to Death)

```{r, echo = T, message = F, warning = F, fig.width = 10, fig.height = 7}
# Some covariates of interest
fit_er <- survfit(Surv(LMD.surv, Status) ~ ER, data = LMD_cohort)
fit_pr <- survfit(Surv(LMD.surv, Status) ~ PR, data = LMD_cohort)
fit_her2 <- survfit(Surv(LMD.surv, Status) ~ HER2, data = LMD_cohort)

ggsurvplot(
  fit_er,
  data = LMD_cohort,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Positive", "Negative"),  # Change legend labels
  legend.title = 'ER Status',
  risk.table.height = 0.2, # Useful to change when you have multiple groups
  ggtheme = theme_bw(),      # Change ggplot2 theme
  font.title = c(22, "bold"),
  font.subtitle = c(21, "bold.italic"),
  font.caption = c(19, "plain"),
  font.x = c(19, "bold.italic"),
  font.y = c(19, "bold.italic"),
  font.tickslab = c(18, "plain"),
  font.legend = c(18, "plain"),
  fontsize = 6
)



ggsurvplot(
  fit_pr,
  data = LMD_cohort,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Positive", "Negative"),    # Change legend labels
  legend.title = 'PR Status',
  risk.table.height = 0.2, # Useful to change when you have multiple groups
  ggtheme = theme_bw(),      # Change ggplot2 theme
  font.title = c(22, "bold"),
  font.subtitle = c(21, "bold.italic"),
  font.caption = c(19, "plain"),
  font.x = c(19, "bold.italic"),
  font.y = c(19, "bold.italic"),
  font.tickslab = c(18, "plain"),
  font.legend = c(18, "plain"),
  fontsize = 6
)

ggsurvplot(
  fit_her2,
  data = LMD_cohort,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Positive", "Negative"),    # Change legend labels
  legend.title = 'HER2 Status',
  risk.table.height = 0.2, # Useful to change when you have multiple groups
  ggtheme = theme_bw(),      # Change ggplot2 theme
  font.title = c(22, "bold"),
  font.subtitle = c(21, "bold.italic"),
  font.caption = c(19, "plain"),
  font.x = c(19, "bold.italic"),
  font.y = c(19, "bold.italic"),
  font.tickslab = c(18, "plain"),
  font.legend = c(18, "plain"),
  fontsize = 6
)
```

### 8.a Linear Regression for primary objective 1.a

```{r, echo = T, message = F, warning = F, fig.asp = 0.75}



lm1 <- with(base_LMD, lm(Brc.LMD ~ ER + PR + HER2 + Age + Race + Stage))


lm1.1 <- with(base_LMD, lm(Brc.LMD ~ ER + Age + Race + Stage))


lm1.2 <- with(base_LMD, lm(Brc.LMD ~ PR + Age + Race + Stage))



lm1.3 <- with(base_LMD, lm(Brc.LMD ~ HER2 + Age + Race + Stage))


```

#### Summary of regression results (n = 139)

```{r, echo = T, message = F, warning = F}
get_pool_coef(lm1) %>% 
  full_join(get_pool_coef(lm1.1), by = 'Variables') %>% 
  full_join(get_pool_coef(lm1.2), by = 'Variables') %>% 
  full_join(get_pool_coef(lm1.3), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 4)), format = 'html', align = rep('r', 20), 
      caption = 'Table.1a Linear Regression: time from initial breast cancer to LMD for baseline HRs') %>% 
  add_header_above(c(" " = 1, "ER + PR + HER2 + Age + Race + Stage" = 2, "ER + Age + Race + Stage" = 2, 
                     "PR + Age + Race + Stage" = 2, "HER2 + Age + Race + Stage" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```


-  ER (model 2) and PR (model 3) confirmed our hypotheses whereas it is not applicable to HER2.
-   Sample interpretation 1: Presents of ER- is associated with a shorter time of 977.5 (p = 0.004) days from initial breast cancer to LMD adjusting for age, race and stage, as compared to ER+.
-   Sample interpretation 2: Presents of PR- is associated with a shorter time of 948 (p = 0.003) days from initial breast cancer to LMD adjusting for age, race and stage, as compared to PR+.

<!-- <!-- It is likely for ER and PR to be correlated because only PR is significant when including both ER and PR in a linear model, whereas ER and PR achieve significance when entering the model separately. Thus, it would be helpful to assess the correlation between ER and PR. --> -->

<!-- # ```{r, echo = T, message = F, warning = F} -->
<!-- # cor.test(as.numeric(LMD_cohort$ER), as.numeric(LMD_cohort$PR)) -->
<!-- # ``` -->
<!-- # - A $\rho$ of 0.75 indicates a strong correlation -->
<!-- # - The test result shows the value of $\rho$ is significantly different from 0. -->


### 8.b Sensitity analysis: include only patients whose 1st metastasis was non-brain metastasis (not included).

```{r, echo = T, message = F, warning = F, fig.asp = 0.75}



lm1_subgroup <- with(base_LMD, lm(Brc.LMD ~ ER*First.met.site + PR + HER2 + Age + Race + Stage))


lm1.1_subgroup <- with(base_LMD, lm(Brc.LMD ~ ER*First.met.site + Age + Race + Stage))

summary(pool(lm1.1_subgroup))

test <- emmeans(lm1.1_subgroup, ~ ER * First.met.site)

contrast(test)


lm1.2_subgroup <- with(base_LMD, lm(Brc.LMD ~ PR*First.met.site + Age + Race + Stage))



lm1.3_subgroup <- with(base_LMD, lm(Brc.LMD ~ HER2*First.met.site + Age + Race + Stage))





```

#### Summary of regression results (n = 139)

```{r, echo = T, message = F, warning = F, results = 'asis'}
get_pool_coef(lm1_subgroup) %>% 
  full_join(get_pool_coef(lm1.1_subgroup), by = 'Variables') %>% 
  full_join(get_pool_coef(lm1.2_subgroup), by = 'Variables') %>% 
  full_join(get_pool_coef(lm1.3_subgroup), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 4)), format = 'html', align = rep('r', 20),
      caption = 'Table.1b Linear Regression: time from initial breast cancer to LMD by 1st met location (Non brain vs. Brain) for baseline HRs') %>% 
  add_header_above(c(" " = 1, "ER * 1st met location + PR + HER2 + Age + Race + Stage" = 2, "ER * 1st met location + Age + Race + Stage" = 2, 
                     "PR * 1st met location + Age + Race + Stage" = 2, "HER2 * 1st met location + Age + Race + Stage" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```

-   In full model, only stage 4 is significant.
-   Looking at ER, PR and HER2 one by one, the main effect of 1st met location was not significant and the interaction between 1st met location and each HR was not significant. Therefore, it's not meaningful to include the 1st met location and interaction between 1st met location and HR.


### 9. Cox Regression for primary objective 1.b

```{r, echo = T, message = F, warning = F, fig.asp = 0.75}


cox.base.lmd.1 <- with(base_LMD, coxph(Surv(LMD.surv, Status) ~ ER + PR + HER2 + Age + Race + Stage))


cox.base.lmd.2 <- with(base_LMD, coxph(Surv(LMD.surv, Status) ~ ER + Age + Race + Stage))


cox.base.lmd.3 <- with(base_LMD, coxph(Surv(LMD.surv, Status) ~ PR + Age + Race + Stage))


cox.base.lmd.4 <- with(base_LMD, coxph(Surv(LMD.surv, Status) ~ HER2 + Age + Race + Stage))


cox.base.lmd.5 <- with(base_LMD, coxph(Surv(LMD.surv, Status) ~ ER + HER2 + Age + Race + Stage))


cox.base.lmd.6 <- with(base_LMD, coxph(Surv(LMD.surv, Status) ~ PR + HER2 + Age + Race + Stage))


```

#### Summary of Cox regression results (n = 139)

```{r, echo = T, message = F, warning = F, results = 'asis'}
get_pool_coef_cox(cox.base.lmd.1) %>% 
  full_join(get_pool_coef_cox(cox.base.lmd.2), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.base.lmd.3), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.base.lmd.4), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.base.lmd.5), by = 'Variables') %>%
  full_join(get_pool_coef_cox(cox.base.lmd.6), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 6)), format = 'html', align = rep('r', 20),
      caption = 'Table.3 Cox Regression: time from LMD to death for baseline HRs') %>% 
  add_header_above(c(" " = 1, "ER + PR + HER2 + Age + Race + Stage" = 2, "ER + Age + Race + Stage" = 2, 
                     "PR + Age + Race + Stage" = 2, "HER2 + Age + Race + Stage" = 2, "ER + HER2 + Age + Race + Stage" = 2,
                     "PR + HER2 + Age + Race + Stage" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```

-   In full model, only stage 2 is significant.
-   Looking at ER, PR and HER2 one by one, the hazard ratio of ER and PR confirmed our hypotheses whereas it is not applicable to HER2.
-   Sample interpretation 1: The present of ER- is associated with a 70% increased risk of death adjusting for age, race and stage, as compared to ER+.
-   Sample interpretation 2: The present of PR- is associated with a 70% increased risk of death adjusting for age, race, stage and HER2, as compared to PR+.



### 10. Two by Two table: ER by 1st metastasis site (Non-brain vs. Brain); PR by 1st metastasis site (Non-brain vs. Brain)

```{r, echo = T, message = F, warning = F}
table(LMD_cohort$First.met.bi, factor(LMD_cohort$ER)) %>%
  kbl(align = 'ccc', row.names = T, caption = 'ER: Non-brain Met vs. Brain Met (excluding LMD)') %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)

chisq.test(table(LMD_cohort$First.met.bi, LMD_cohort$ER), correct = T)

table(LMD_cohort$First.met.bi, LMD_cohort$PR) %>%
  kbl(align = 'ccc', row.names = T, caption = 'PR: Non-brain Met vs. Brain Met (excluding LMD)') %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)

chisq.test(table(LMD_cohort$First.met.bi, LMD_cohort$PR), correct = T)


table(LMD_cohort$First.met.bi.LMD, factor(LMD_cohort$ER)) %>%
  kbl(align = 'ccc', row.names = T, caption = 'ER: Non-brain Met vs. Brain Met (including LMD)') %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)

chisq.test(table(LMD_cohort$First.met.bi.LMD, LMD_cohort$ER), correct = T)

table(LMD_cohort$First.met.bi.LMD, LMD_cohort$PR) %>%
  kbl(align = 'ccc', row.names = T, caption = 'PR: Non-brain Met vs. Brain Met (including LMD)') %>%
  kable_classic(full_width = F, html_font = "Arial", position = "left", font_size = 14)

chisq.test(table(LMD_cohort$First.met.bi.LMD, LMD_cohort$PR), correct = T)


```

### 11. Hormone Discordance
### i.ER (Overall Sample: n = 130)
```{r, echo = T, message = F, warning = F}
library(rms)
library(pec)
library(prodlim)


# ER
cox.1st.base.non.brain1 <- with(base_1st_met, coxph(Surv(Brc.first.met, Event == 1) ~ ER + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))



cox.1st.base.brain1 <- with(base_1st_met, coxph(Surv(Brc.first.met, Event == 2) ~ ER + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))



cox.1st.HD.non.brain1 <- with(HD_1st_met, coxph(Surv(Tstart, Tstop, Event == 1) ~ ER + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))



cox.1st.HD.brain1 <- with(HD_1st_met, coxph(Surv(Tstart, Tstop, Event == 2) ~ ER + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))


```


```{r, echo = T, message = F, warning = F, fig.width = 10, fig.height = 7}
get_pool_coef_cox(cox.1st.base.non.brain1) %>% 
  full_join(get_pool_coef_cox(cox.1st.base.brain1), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.1st.HD.non.brain1), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.1st.HD.brain1), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 4)), format = 'html', align = rep('r', 20),
      caption = 'Table.4a Cox Regression: time from initial breast cancer to 1st met for ER') %>% 
  add_header_above(c(" " = 1, "Baseline ER: 1st met at Non-brain site" = 2, "Baseline ER: 1st met at brain site" = 2, 
                     "Time-dependent ER: 1st met at Non-brain site" = 2, "Time-dependent ER: 1st met at brain site" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```
#### Summary
- Stage 4 is significant across all models.
- Baseline and time-dependent ER is significant in the model examining time to 1st metastasis as brain metastasis.
- The result of the model with baseline ER demonstrated that the ER- is associated with higher risk of 1st metastasis in brain, adjusting for race, age and stage, at any possible time before the event, even if there is 1 or multiple switch.
- Sample interpretation: The present of ER- is associated with a 140% increased risk of 1st metastasis in brain, adjusting for age, race and stage, as compared to ER+.
- The result of the model with time-dependent ER demonstrated that the ER- is associated with higher risk of 1st metastasis in brain, adjusting for race, age and stage, at any possible time before the event, even if there is 1 or multiple switch.
- Sample interpretation: After considering the ER status over time, the present of ER- is associated with a 220% increased risk of 1st metastasis in brain, adjusting for age, race and stage, as compared to ER+.



### ii.PR (Overall Sample: n = 130)
```{r, echo = T, message = F, warning = F}

# PR
cox.1st.base.non.brain2 <- with(base_1st_met, coxph(Surv(Brc.first.met, Event == 1) ~ PR + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))



cox.1st.base.brain2 <- with(base_1st_met, coxph(Surv(Brc.first.met, Event == 2) ~ PR + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))



cox.1st.HD.non.brain2 <- with(HD_1st_met, coxph(Surv(Tstart, Tstop, Event == 1) ~ PR + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))



cox.1st.HD.brain2 <- with(HD_1st_met, coxph(Surv(Tstart, Tstop, Event == 2) ~ PR + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))


```

```{r, echo = T, message = F, warning = F, fig.width = 10, fig.height = 7}
get_pool_coef_cox(cox.1st.base.non.brain2) %>% 
  full_join(get_pool_coef_cox(cox.1st.base.brain2), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.1st.HD.non.brain2), by = 'Variables') %>% 
  full_join(get_pool_coef_cox(cox.1st.HD.brain2), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 4)), format = 'html', align = rep('r', 20),
      caption = 'Table.4b Cox Regression: time from initial breast cancer to 1st met for PR') %>% 
  add_header_above(c(" " = 1, "Baseline PR: 1st met at Non-brain site" = 2, "Baseline PR: 1st met at brain site" = 2, 
                     "Time-dependent PR: 1st met at Non-brain site" = 2, "Time-dependent PR: 1st met at brain site" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```
#### Summary
- Stage 4 is significant across all models.
- Baseline and time-dependent PR is significant in the model examining time to 1st metastasis as brain metastasis.
- The result of the model with baseline PR demonstrated that the PR- is associated with higher risk of 1st metastasis in brain, adjusting for race, age and stage, at any possible time before the event, even if there is 1 or multiple switch.
- Sample interpretation: The present of PR- is associated with a 110% increased risk of 1st metastasis in brain, adjusting for age, race and stage, as compared to ER+.
- The result of the model with time-dependent ER demonstrated that the PR- is associated with higher risk of 1st metastasis in brain, adjusting for race, age and stage, at any possible time before the event, even if there is 1 or multiple switch.
- Sample interpretation: After considering the PR status over time, the present of PR- is associated with a 180% increased risk of 1st metastasis in brain, adjusting for age, race and stage, as compared to PR+.



### iii. Additional analysis (Outcome = Time from Brc to brain metastsis or LMD, subgroups = patients who had 1st metastatsis as non-brain metastasis or brain metastasis)

### KM plots by ER, PR and HER2 status (Brc to LMD)

```{r, echo = T, message = F, warning = F, fig.width = 10, fig.height = 7}
# Some covariates of interest

LMD_cohort <- LMD_cohort %>% mutate(Event = 1)

fit_er <- survfit(Surv(Brc.LMD, Event) ~ ER, data = LMD_cohort)
fit_pr <- survfit(Surv(Brc.LMD, Event) ~ PR, data = LMD_cohort)
fit_her2 <- survfit(Surv(Brc.LMD, Event) ~ HER2, data = LMD_cohort)

ggsurvplot(
  fit_er,
  data = LMD_cohort,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Positive", "Negative"),  # Change legend labels
  legend.title = 'ER Status',
  risk.table.height = 0.2, # Useful to change when you have multiple groups
  ggtheme = theme_bw(),      # Change ggplot2 theme
  font.title = c(22, "bold"),
  font.subtitle = c(21, "bold.italic"),
  font.caption = c(19, "plain"),
  font.x = c(19, "bold.italic"),
  font.y = c(19, "bold.italic"),
  font.tickslab = c(18, "plain"),
  font.legend = c(18, "plain"),
  fontsize = 6
)

ggsurvplot(
  fit_pr,
  data = LMD_cohort,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Positive", "Negative"),    # Change legend labels
  legend.title = 'PR Status',
  risk.table.height = 0.2, # Useful to change when you have multiple groups
  ggtheme = theme_bw(),      # Change ggplot2 theme
  font.title = c(22, "bold"),
  font.subtitle = c(21, "bold.italic"),
  font.caption = c(19, "plain"),
  font.x = c(19, "bold.italic"),
  font.y = c(19, "bold.italic"),
  font.tickslab = c(18, "plain"),
  font.legend = c(18, "plain"),
  fontsize = 6
)

ggsurvplot(
  fit_her2,
  data = LMD_cohort,
  size = 1,                 # change line size
  palette =
    c("#E7B800", "#2E9FDF"),# custom color palettes
  conf.int = TRUE,          # Add confidence interval
  pval = TRUE,              # Add p-value
  risk.table = TRUE,        # Add risk table
  risk.table.col = "strata",# Risk table color by groups
  legend.labs =
    c("Positive", "Negative"),    # Change legend labels
  legend.title = 'HER2 Status',
  risk.table.height = 0.2, # Useful to change when you have multiple groups
  ggtheme = theme_bw(),      # Change ggplot2 theme
  font.title = c(22, "bold"),
  font.subtitle = c(21, "bold.italic"),
  font.caption = c(19, "plain"),
  font.x = c(19, "bold.italic"),
  font.y = c(19, "bold.italic"),
  font.tickslab = c(18, "plain"),
  font.legend = c(18, "plain"),
  fontsize = 6
)
```


### ER (n = 137)
```{r, echo = T, message = F, warning = F}

# ER
cox.base.lmd <- with(base_HD_LMD, coxph(Surv(Brc.LMD, Event) ~ ER + First.met.brain + First.met.non.brain + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))




cox.HD.lmd <- with(HD_LMD, coxph(Surv(Tstart, Tstop, Event) ~ ER + First.met.brain + First.met.non.brain + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))


```

```{r, echo = T, message = F, warning = F, fig.width = 10, fig.height = 7}

get_pool_coef_cox(cox.base.lmd) %>% 
  full_join(get_pool_coef_cox(cox.HD.lmd), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 2)), format = 'html', align = rep('r', 20),
      caption = 'Table.5a Cox Regression: time from initial breast cancer to LMD by 1st met location (Non brain vs. Brain) for ER') %>% 
  add_header_above(c(" " = 1, "Baseline ER + Baseline 1st met in brain + Baseline 1st met in non-brain + Race + Age + Stage" = 2, 
                     "Time-dependent ER + Time-dependent 1st met in brain + Time-dependent 1st met in non-brain + Race + Age + Stage" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```

#### Summary
- ER is significant in both model.
- 1st met in brain is significant in both model, while 1st met in non-brain is significant only in mdoel with time-dependent ER and 1st met location.
- HR decreased when including ER as time-dependent.
- Sample interpretation 1: Having 1st met in brain is associated with a 140% increased in hazard of LMD, adjusting for age, race, stage and ER, as compared to not having any 1st met yet at any given time point.
- Sample interpretation 2: Having 1st met in non-brain is associated with a 60% increased in hazard of LMD, adjusting for age, race, stage and ER, as compared to not having any 1st met yet at any given time point.


### PR (n = 137)
```{r, echo = T, message = F, warning = F}


# PR
cox.base.lmd <- with(base_HD_LMD, coxph(Surv(Brc.LMD, Event) ~ PR + First.met.brain + First.met.non.brain + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))

cox.HD.lmd <- with(HD_LMD, coxph(Surv(Tstart, Tstop, Event) ~ PR + First.met.brain + First.met.non.brain + Race + Age + Stage + cluster(ID), 
                                         na.action = "na.omit", id = ID))




```

```{r, echo = T, message = F, warning = F, fig.width = 10, fig.height = 7}

get_pool_coef_cox(cox.base.lmd) %>% 
  full_join(get_pool_coef_cox(cox.HD.lmd), by = 'Variables') %>% 
  kbl(col.names = c("Variable", rep(c('Coefficients [95% CI]', 'p-value'), 2)), format = 'html', align = rep('r', 20),
      caption = 'Table.5b Cox Regression: time from initial breast cancer to LMD by 1st met location (Non brain vs. Brain) for PR') %>% 
  add_header_above(c(" " = 1, "Baseline PR + Baseline 1st met in brain + Baseline 1st met in non-brain + Race + Age + Stage" = 2, 
                     "Time-dependent PR + Time-dependent 1st met in brain + Time-dependent 1st met in non-brain + Race + Age + Stage" = 2)) %>% 
  kable_paper(bootstrap_options = "striped", full_width = T)
```

#### Summary
- PR is significant in both model.
- 1st met in brain is significant in both model, while 1st met in non-brain is significant only in mdoel with time-dependent PR and 1st met location.
- HR decreased when including PR as time-dependent.
- Sample interpretation 1: Having 1st met in brain is associated with a 140% increased in hazard of LMD, adjusting for age, race, stage and PR, as compared to not having the any 1st met at any given time point.
- Sample interpretation 2: Having 1st met in non-brain is associated with a 50% increased in hazard of LMD, adjusting for age, race, stage and PR, as compared to not having any 1st met at any given time point.

